---
title: "Testing Lahman 6.0"
author: "Martin Monkman"
date: "2017-07-17"
output: html_notebook
---

This markdown document incorporates a variety of short scripts that draw on the various tables in the `Lahman` package. (See the Lahman project page on RForge for more details <http://lahman.r-forge.r-project.org/>.)

Note that some of scripts appear in the documentation of other R packages; in those cases, the original source is noted prior to the script.

First of all, install and load the new version of `Lahman`, and load the `tidyverse` package.

```{r}
# install.packages("Lahman", repos="http://R-Forge.R-project.org")

devtools::install_github("cdalzell/Lahman", ref="develop")
library(Lahman)

library(tidyverse)
#
```


## Master & Fielding

This uses an inner join to merge the two tables `Master` and `Fielding` into a new data frame "MasterFielding".  The first approach relies on the `merge` instruction, while the second uses the much faster `dplyr::inner_join` command.


```{r}
# throwing by position
# version 1 - "merge"
MasterFielding <- data.frame(merge(Master, Fielding, by="playerID"))
MasterFielding <- merge(Master, Fielding, by="playerID")
system.time(MasterFielding <- merge(Master, Fielding, by="playerID"))
# the dplyr version -- faster
MasterFielding <- inner_join(Fielding, Master, by="playerID")
system.time(MasterFielding <- inner_join(Fielding, Master, by="playerID"))
#
#
# a count of games played, by position
MasterFielding <- subset(MasterFielding, POS != "OF" & yearID > "1944")
#
MasterFielding %>%
  group_by(playerID, POS, throws) %>%
  summarise(gamecount = sum(G)) %>%
  arrange(desc(gamecount)) %>% 
  head(5)
#
MasterFielding <- inner_join(Fielding, Master, by="playerID")
# select only those seasons since 1945 and 
# omit the records that are OF summary (i.e. leave the RF, CF, and LF)
MasterFielding <- subset(MasterFielding, POS != "OF" & yearID > "1944")
#
```

Now with the MasterFielding table, we can use the `dplyr::summarise` command to generate a new table, "Player_games", that in turn is summarised into a table counting the games played at each position.

```{r}

Player_games <- MasterFielding %>%
  group_by(playerID, POS, throws) %>%
  summarise(gamecount = sum(G))
Player_POS <- Player_games %>%
  group_by(POS, throws) %>%
  summarise(playercount = length(gamecount))
head(Player_POS)
#

```

## Batting (dplyr reference)

This chunk of code is lifted straight from the "Data transformation" chapter of the online version of the book [_R for Data Science_](http://r4ds.had.co.nz/transform.html) by Garret Grolemund and Hadley Wickham (Specifically, section 5.6.3.


```{r}

# Convert to a tibble so it prints nicely
batting <- as_tibble(Lahman::Batting)

batters <- batting %>% 
  group_by(playerID) %>% 
  summarise(
    ba = sum(H, na.rm = TRUE) / sum(AB, na.rm = TRUE),
    ab = sum(AB, na.rm = TRUE)
  )

batters %>% 
  filter(ab > 100) %>% 
  ggplot(mapping = aes(x = ab, y = ba)) +
    geom_point() + 
    geom_smooth(se = FALSE)
#> `geom_smooth()` using method = 'gam'
```

A second chunk from _R for Data Science_ (same section).

```{r}

batters %>% 
  arrange(desc(ba))
#> # A tibble: 18,659 × 3
#>    playerID    ba    ab
#>       <chr> <dbl> <int>
#> 1 abramge01     1     1
#> 2 banisje01     1     1
#> 3 bartocl01     1     1
#> 4  bassdo01     1     1
#> 5 birasst01     1     2
#> 6 bruneju01     1     1
#> # ... with 1.865e+04 more rows

```




### `dplyr` vignettes

[Window functions](http://dplyr.tidyverse.org/articles/window-functions.html)

```{r}
#1
batting <- Lahman::Batting %>%
  as_tibble() %>%
  select(playerID, yearID, teamID, G, AB:H) %>%
  arrange(playerID, yearID, teamID) %>%
  semi_join(Lahman::AwardsPlayers, by = "playerID")

players <- batting %>% group_by(playerID)


#2
# For each player, find the two years with most hits
filter(players, min_rank(desc(H)) <= 2 & H > 0)
# Within each player, rank each year by the number of games played
mutate(players, G_rank = min_rank(G))

# For each player, find every year that was better than the previous year
filter(players, G > lag(G))
# For each player, compute avg change in games played per year
mutate(players, G_change = (G - lag(G)) / (yearID - lag(yearID)))

# For each player, find all where they played more games than average
filter(players, G > mean(G))
# For each, player compute a z score based on number of games played
mutate(players, G_z = (G - mean(G)) / sd(G))

```


ranking functions


```{r}

#1
filter(players, cume_dist(desc(G)) < 0.1)

#2
by_team_player <- group_by(batting, teamID, playerID)
by_team <- summarise(by_team_player, G = sum(G))
by_team_quartile <- group_by(by_team, quartile = ntile(G, 4))
summarise(by_team_quartile, mean(G))

```


lead and lag

```{r}

#1
# Compute the relative change in games played
mutate(players, G_delta = G - lag(G))

#2
# Find when a player changed teams
filter(players, teamID != lag(teamID))






```



aggregates

```{r}

# cumulative

filter(players, cumany(G > 150))


# recycled

filter(players, G > mean(G))
filter(players, G < median(G))

filter(players, ntile(G, 2) == 2)

mutate(players, career_year = yearID - min(yearID) + 1)


mutate(players, G_z = (G - mean(G)) / sd(G))


```



### Steven Jay Gould's graph

From Jim Albert's 2015-08-31 blog post, ["Redrawing Steven Jay Gould's Graph"](https://baseballwithr.wordpress.com/category/lahman/).

```{r}

# 1
S <- summarize(group_by(Batting, yearID, playerID),
                        H=sum(H), AB=sum(AB),
                        BB=sum(BB), HBP=sum(HBP, na.rm=TRUE),
                        SF=sum(SF, na.rm=TRUE))

#2
ST <- summarize(group_by(Teams, yearID),
               Games = round(mean(W + L)),
               min.AB = 3 * Games)
S2 <- merge(S, ST, by="yearID")
S.regular <- filter(S2, AB >= min.AB) 

#3
SD.data <- summarize(group_by(S.regular, yearID),
                     SD.AVG = sd(H / AB),
                     SD.OBP = sd((H + BB + HBP) / (AB + BB + HBP + SF)))


#4
ggplot(SD.data, aes(SD.AVG, yearID)) + geom_point() +
  xlim(.005, .075) + 
  ggtitle("Standard Deviations of AVG of Regulars (Gould's Display)")

#5
ggplot(SD.data, aes(yearID, SD.AVG)) + geom_point() + 
  geom_smooth(se=FALSE, span=0.35) + 
  ggtitle("Standard Deviations of AVG of Regulars")

#6
ggplot(SD.data, aes(yearID, SD.OBP)) + 
  geom_point() + geom_smooth(se=FALSE, span=0.35) +
  ggtitle("Standard Deviations of OBP of Regulars")



```







---

## Reference list
[Lahman at CRAN](http://cran.r-project.org/web/packages/Lahman/index.html)

[Lahman at RForge](http://lahman.r-forge.r-project.org/)

### dplyr

[dplyr at CRAN](http://cran.r-project.org/web/packages/dplyr/index.html)

[dplyr: Introduction to dplyr (at CRAN)](http://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html)

[dplyr: Vignettes](http://cran.r-project.org/web/packages/dplyr/vignettes/databases.html)