---
title: "Runs per game by MLB team"
author: "Martin Monkman"
date: "Sunday, November 09, 2014"
output: html_document
---

## RETRACING MY STEPS

### REFERENCE MATERIALS 

#### SHINY

[RStudio Shiny widget gallery](http://shiny.rstudio.com/gallery/widget-gallery.html)

[RStudio Shiny function reference](http://shiny.rstudio.com/reference/shiny/latest/)



### CODE CHUNK #1 
original source: [Major League Baseball run scoring trends with R's Lahman package](http://bayesball.blogspot.ca/2013/06/major-league-baseball-run-scoring.html)
gist: https://gist.github.com/MonkmanMH/5711584

```{r}
# load the necessary packages and data set "Teams"
#install.packages("Lahman")
library(dplyr)
library(ggplot2)
library(Lahman)
#
# load the Lahman data table "Teams"
data(Teams)
#
#
# CREATE LEAGUE SUMMARY TABLES
# ============================
#
# select a sub-set of teams from 1901 [the establishment of the American League] forward to 2012
Teams_sub <- as.data.frame(subset (Teams, yearID > 1900 & lgID != "FL"))
#
# calculate each team's average runs and runs allowed per game
Teams_sub$RPG <- Teams_sub$R / Teams_sub$G
Teams_sub$RAPG <- Teams_sub$RA / Teams_sub$G
#
# create new data frame with season totals for each league
LG_RPG <- aggregate(cbind(R, RA, G) ~ yearID + lgID, data = Teams_sub, sum)
# calculate league + season runs and runs allowed per game
LG_RPG$LG_RPG <- LG_RPG$R / LG_RPG$G
LG_RPG$LG_RAPG <- LG_RPG$RA / LG_RPG$G
 
head(LG_RPG)
#
```
 
An alternative approach to creating the LG_RPG data frame, using dplyr. The only difference (in this implementation) is that the original is sorted by league, whereas the dplyr is sorted by year.
 
```{r}
 
# dplyr approach to creating LG_RPG
LG_RPG <- Teams %>%
  filter(yearID > 1900, lgID != "FL") %>%
  group_by(yearID, lgID) %>%
  summarise(R=sum(R), RA=sum(RA), G=sum(G)) %>%
  mutate(RPG=R/G, RAPG=RA/G)
 
head(LG_RPG)

# and a version with just the MLB totals
MLB_RPG <- Teams %>%
  filter(yearID > 1900, lgID != "FL") %>%
  group_by(yearID) %>%
  summarise(R=sum(R), RA=sum(RA), G=sum(G)) %>%
  mutate(RPG=R/G, RAPG=RA/G)
 
head(MLB_RPG)


```
 
Filtering the LG_RPG file into the separate league tables.  This is no longer necessary, as dplyr and ggplot can filter on the fly.
 
```{r}#
# select a sub-set of teams from 1901 [the establishment of the American League] forward to 2012
# read the data into separate league tables
ALseason <- (subset (LG_RPG, yearID > 1900 & lgID == "AL"))
NLseason <- (subset (LG_RPG, yearID > 1900 & lgID == "NL"))
#
#
# Use dplyr (and the inner_join function) to create Teams.merge
Teams.merge <- inner_join(Teams_sub, LG_RPG, by = c("yearID", "lgID"))
head(Teams.merge)
tail(Teams.merge)
#
#
# +++++++++++++++++++++++++++++++++++++++++++++++++++
#
# RUNS SCORED PER GAME BY LEAGUE
# ==============================
#
# web references:
# http://princeofslides.blogspot.ca/2011/05/sab-r-metrics-basics-of-loess.html
# http://research.stowers-institute.org/efg/R/Statistics/loess.htm
#
# start with American League
# create new object ALRunScore.LO for loess model
ALRunScore.LO <- loess(ALseason$LG_RPG ~ ALseason$yearID)
ALRunScore.LO.predict <- predict(ALRunScore.LO)
#
# create new objects RunScore.Lo.XX for loess models with "span" control
ALRunScore.LO.25 <- loess(ALseason$LG_RPG ~ ALseason$yearID, span=0.25)  
ALRunScore.LO.25.predict <- predict(ALRunScore.LO.25)
#
ALRunScore.LO.5 <- loess(ALseason$LG_RPG ~ ALseason$yearID, span=0.5)	
ALRunScore.LO.5.predict <- predict(ALRunScore.LO.5)
#
# plot the data, add loess curve
ylim <- c(3,6)
plot(ALseason$LG_RPG ~ ALseason$yearID,
ylim = ylim,
main = "American League: runs per team per game, 1901-2012",
xlab = "year", ylab = "runs per game")
# loess predicted value line
lines(ALseason$yearID, ALRunScore.LO.predict, lty="solid", col="red", lwd=2)
lines(ALseason$yearID, ALRunScore.LO.25.predict, lty="dashed", col="blue", lwd=2)
lines(ALseason$yearID, ALRunScore.LO.5.predict, lty="dotdash", col="black", lwd=2)
legend(1980, 3.5,
c("default", "span=0.25", "span=0.50"),
lty=c("solid", "dashed", "dotdash"),
col=c("red", "blue", "black"),
lwd=c(2, 2, 2))
grid()
#
#
# NATIONAL LEAGUE
# create new object RunScore.LO for loess model
NLRunScore.LO <- loess(NLseason$LG_RPG ~ NLseason$yearID)	
NLRunScore.LO.predict <- predict(NLRunScore.LO)
#
# objects with span control in loess model
NLRunScore.LO.25 <- loess(NLseason$LG_RPG ~ NLseason$yearID, span=0.25)	
NLRunScore.LO.25.predict <- predict(NLRunScore.LO.25)
NLRunScore.LO.5 <- loess(NLseason$LG_RPG ~ NLseason$yearID, span=0.5)	
NLRunScore.LO.5.predict <- predict(NLRunScore.LO.5)
#
# plot the data, add loess curve
ylim <- c(3,6)
plot(NLseason$LG_RPG ~ NLseason$yearID,
pch=2, col="black",
ylim = ylim,
main = "National League: runs per team per game, 1901-2012",
xlab = "year", ylab = "runs per game")
# loess predicted value line
lines(NLseason$yearID, NLRunScore.LO.predict, lty="solid", col="blue", lwd=2)
lines(NLseason$yearID, NLRunScore.LO.25.predict, lty="dashed", col="red", lwd=2)
lines(NLseason$yearID, NLRunScore.LO.5.predict, lty="dotdash", col="black", lwd=2)
# chart tidying
legend(1980, 3.5,
c("default", "span=0.25", "span=0.50"),
lty=c("solid", "dashed", "dotdash"),
col=c("blue", "red", "black"),
lwd=c(2, 2, 2))
grid()
#
#
# MULTI-PLOT -- MERGING AL AND NL RESULTS
# plot individual years as lines
ylim <- c(3,6)
# start with AL line
plot(ALseason$LG_RPG ~ ALseason$yearID,
type="l", lty="solid", col="red", lwd=2,
main = "Runs per team per game, 1901-2012",
ylim = ylim,
xlab = "year", ylab = "runs per game")
# add NL line
lines(NLseason$yearID, NLseason$LG_RPG, lty="solid", col="blue", lwd=2)
# chart additions
grid()
legend(1900, 3.5, c("AL", "NL"), lty=c("solid", "solid"), col=c("red", "blue"), lwd=c(2, 2))
#
#
# plot loess curves (span=0.25)
ylim <- c(3,6)
# start with AL line
plot(ALRunScore.LO.25.predict ~ ALseason$yearID,
type="l", lty="solid", col="red", lwd=2,
main = "Runs per team per game, 1901-2012",
ylim = ylim,
xlab = "year", ylab = "runs per game")
# add NL line
lines(NLseason$yearID, NLRunScore.LO.25.predict, lty="dashed", col="blue", lwd=2)
# chart additions
legend(1900, 3.5,
c("AL (span=0.25)", "NL (span=0.25)"),
lty=c("solid", "dashed"),
col=c("red", "blue"),
lwd=c(2, 2))
grid()
#
#
# plot loess curves (span=0.50)
ylim <- c(3,6)
# start with AL line
plot(ALRunScore.LO.5.predict ~ ALseason$yearID,
type="l", lty="solid", col="red", lwd=2,
main = "Runs per team per game, 1901-2012",
ylim = ylim,
xlab = "year", ylab = "runs per game")
# add NL line
lines(NLseason$yearID, NLRunScore.LO.5.predict, lty="dashed", col="blue", lwd=2)
# chart additions
legend(1900, 3.5,
c("AL (span=0.50)", "NL (span=0.50)"),
lty=c("solid", "dashed"),
col=c("red", "blue"),
lwd=c(2, 2))
grid()
#
#
#
# plot multiple loess curves (span=0.50 and 0.25)
ylim <- c(3,6)
# start with AL line
plot(ALRunScore.LO.5.predict ~ ALseason$yearID,
type="l", lty="solid", col="red", lwd=2,
main = "Runs per team per game, 1901-2012",
ylim = ylim,
xlab = "year", ylab = "runs per game")
# add NL line
lines(NLseason$yearID, NLRunScore.LO.5.predict, lty="solid", col="blue", lwd=2)
# add 0.25 lines
lines(ALseason$yearID, ALRunScore.LO.25.predict, lty="dashed", col="red", lwd=2)
lines(NLseason$yearID, NLRunScore.LO.25.predict, lty="dashed", col="blue", lwd=2)
# chart additions
legend(1900, 3.5,
c("AL (span=0.50)", "NL (span=0.50)", "AL (span=0.25)", "NL (span=0.25)"),
lty=c("solid", "solid", "dashed", "dashed"),
col=c("red", "blue", "red", "blue"),
lwd=c(2, 2, 2, 2))
grid()
#
# # # # # # # # # # # # # # # # # #
#
# calculate the difference between the two leagues
# 1. absolute
RunDiff <- (ALseason$LG_RPG - NLseason$LG_RPG)
# 2. LOESS span=0.25
RunDiffLO <- (ALRunScore.LO.25.predict - NLRunScore.LO.25.predict)
#
# plot the LOESS difference
ylim <- c(-1,1)
plot(RunDiffLO ~ ALseason$yearID,
type="l", lty="solid", col="red", lwd=2,
main = "Run scoring trend: AL difference from NL, 1901-2012",
ylim = ylim,
xlab = "year", ylab = "runs per game")
# add line at zero
abline(h = 0, lty="dotdash")
grid()
#
# plot each year difference as line, trend as line
ylim <- c(-1,1.5)
plot(RunDiffLO ~ ALseason$yearID,
type="l", lty="solid", col="red", lwd=3,
main = "Run scoring trend: AL difference from NL, 1901-2012",
ylim = ylim,
xlab = "year", ylab = "runs per game")
# add RunDiff line
lines(ALseason$yearID, RunDiff, lty="solid", col="black", lwd=1)
# add line at zero
abline(h = 0, lty="dotdash")
grid()
#
#
# plot each year difference as bar, trend as line
ylim <- c(-1,1.5)
plot(RunDiff ~ ALseason$yearID,
type="h", lty="solid", col="blue", lwd=2,
main = "Run scoring trend: AL difference from NL, 1901-2012",
ylim = ylim,
xlab = "year", ylab = "runs per game")
# add RunDiff line
lines(ALseason$yearID, RunDiffLO, lty="solid", col="black", lwd=2)
# add line at zero
abline(h = 0, lty="dotdash")
# chart additions
grid()
legend(1900, 1.5,
c("AL difference from NL: absolute", "AL difference from NL, LOESS (span=0.25)"),
lty=c("solid", "solid"),
col=c("blue", "black"),
lwd=c(2, 2))
#
#
```

### CODE CHUNK #1.01 -- ggplot2 
# ggplot version


```{r}
# American League version (requires filtering of LG_RPG into ALseason)
lastyear <- 2112
title <- paste("AL RPG 1901-", lastyear)
ALRPG <- ggplot(ALseason, aes(x=yearID, y=LG_RPG)) +
                geom_point() +
                ggtitle(paste("AL RPG 1901-", lastyear)) +
                xlab("year") + ylab("runs per game")  
ALRPG
#  
ALRPG + stat_smooth(method=loess, span=0.25, level=0.95)
#

# MLB version (uses facets)
lastyear <- 2112
MLBRPG <- ggplot(LG_RPG, aes(x=yearID, y=LG_RPG)) +
                geom_point() +
                ggtitle(paste("MLB RPG 1901-", lastyear)) +
                xlab("year") + ylab("runs per game")  
MLBRPG <- MLBRPG + facet_grid(lgID ~ .)
#  
MLBRPG + stat_smooth(method=loess, span=0.25, level=0.95)
#


```

### CODE CHUNK #2 - TEAM BY TEAM

```{r}


# CREATE INDEX VALUES FOR EACH TEAM
#
# A. Runs per game
#
# create new values to compare the individual team's runs/game compares to the league average that season
# 1. use an index where 100=the league average for that season
Teams.merge$R_index <- Teams.merge$RPG / Teams.merge$LG_RPG * 100
# 2. and Z scores of the index scores
R_index.sd <- sd(Teams.merge$R_index)
R_index.sd
Teams.merge$R_Z <- (Teams.merge$R_index - 100)/R_index.sd
#
# calculate minimum, maximum, and standard deviation
min(Teams.merge$R_index)
max(Teams.merge$R_index)
sd(Teams.merge$R_index)
#
# B. Runs allowed per game
#
# create new values to compare the individual team's runs allowed/game compares to the league average that season
# 1. use an index where 100=the league average for that season
Teams.merge$RA_index <- Teams.merge$RAPG / Teams.merge$LG_RAPG * 100
# 2. and Z scores of the index scores
RA_index.sd <- sd(Teams.merge$RA_index)
RA_index.sd
Teams.merge$RA_Z <- (Teams.merge$RA_index - 100)/R_index.sd
#
# calculate minimum, maximum, and standard deviation
min(Teams.merge$RA_index)
max(Teams.merge$RA_index)
sd(Teams.merge$RA_index)
#
#
# RANK AND SORT BY R_INDEX & RA_INDEX
# 1. low to high (default) 
#   a. rank
Teams.merge$R_index_rank <- rank(Teams.merge$R_index)
Teams.merge$RA_index_rank <- rank(Teams.merge$RA_index)
#   b. Sort
Teams.merge.sort <- Teams.merge[c("yearID","lgID","franchID","R_index", "R_index_rank")] 
Teams.merge.sort <- Teams.merge.sort[order(Teams.merge.sort$R_index),]
Teams.low_off <- as.data.frame (subset(Teams.merge.sort, R_index < 80))
Teams.low_off
# write.csv(Teams.low_off, file="Teams.low_off.csv")
#
# 2. high to low 
#   a. rank (note use of "-" in front of variable name)
Teams.merge$R_index_rank <- rank(-Teams.merge$R_index)
Teams.merge$RA_index_rank <- rank(-Teams.merge$RA_index)
#   b. sort (note use of "decreasing=TRUE" in "order" command)
Teams.merge.sort <- Teams.merge.sort[order(Teams.merge.sort$R_index,decreasing=TRUE),]
Teams.hi_off <- as.data.frame (subset(Teams.merge.sort, R_index > 120))
Teams.hi_off
# write.csv(Teams.hi_off, file="Teams.hi_off.csv")
#
# PLOT!
# index (basic)
hist(Teams.merge$R_index, 
  main="MLB teams 1947-2012: Distribution of scoring",
  xla="Index value (100=league average)")
#
# index with density curve
hist(Teams.merge$R_index, 
  prob=T,
  main="MLB teams 1947-2012: Distribution & density curve of scoring",
  xla="Index value (100=league average)")
lines(density(Teams.merge$R_index))
#
# Z scores
hist(Teams.merge$R_Z, prob=T)
lines(density(Teams.merge$R_Z))
#
#
# and write a new file with the Teams.merge data 
#write.csv(Teams.merge, file="Teams.merge.csv")
#
```

### CODE CHUNK 3 -- TEAMS

```{r}

# THE HISTORICAL RECORD - RUNS PER GAME BY TEAM
#
# public Gist of this code can be found at
# https://gist.github.com/MonkmanMH/4720641
#
# data source: Lahman Database
# http://www.seanlahman.com/baseball-archive/statistics/
# 2012 version (1871-2012)
# data table "Teams.csv"
#
# note: requires that code found at
# https://gist.github.com/4702915
# has already been run, creating the data frame "Teams.merge"
#
# select the team you want from the franchID variable in the Teams.merge data frame
# note the use of double "=" !
Team1 <- as.data.frame(subset (Teams.merge, franchID == "SEA"))
#
# then pull first and last years of the analysis and team name fields
firstyear <- Team1$yearID[1]
lastyear <- tail(Team1$yearID, 1)
#
# either one of these lines of code work to pull the team name
team.name <- Team1$name[1]
team.name <- tail(Team1$name, 1)
# now put the team name, years, and the rest of the title text into a single string
title.text <- paste(team.name, firstyear, "-", lastyear, 
   ": Runs scored relative to the league average")
#
# BASIC PLOT
# index
plot(Team1$R_index ~ Team1$yearID,  
  main=title.text,
  type = "l")
#
#
# TREND LINE -- using loess modeling
# references: 
#  http://princeofslides.blogspot.ca/2011/05/sab-r-metrics-basics-of-loess.html 
#  http://research.stowers-institute.org/efg/R/Statistics/loess.htm
#
# create new object Runs.LO for loess model
RunIndex.LO <- loess(Team1$R_index ~ Team1$yearID)    		
RunIndex.LO.predict <- predict(RunIndex.LO)
#
# plot the data, add loess curve
ylim <- c(60,140)
plot(Team1$R_index ~ Team1$yearID, 
  ylim = ylim, 
  main = title.text, 
  xlab = "year", ylab = "index (league average = 100)")
# chart tidying
  grid()
# loess predicted value line
  lines(Team1$yearID, RunIndex.LO.predict, 
    lty="solid", col="red", lwd=2)	
#
# MODEL VERSION 2 (with more nuance/zig-zag to the trend line)
# create new object RunIndex.LO for loess model, span=0.25 and span=0.5
RunIndex.LO.25 <- loess(Team1$R_index ~ Team1$yearID, span=0.25)				
RunIndex.LO.25.predict <- predict(RunIndex.LO.25)
#
RunIndex.LO.5 <- loess(Team1$R_index ~ Team1$yearID, span=0.5)				
RunIndex.LO.5.predict <- predict(RunIndex.LO.5)
#
# plot the data, add loess curves
ylim <- c(60,140)
plot(Team1$R_index ~ Team1$yearID, 
  ylim = ylim, 
  main = title.text, 
  xlab = "Year", ylab = "Index (league average = 100)")
# loess predicted value line
  lines(Team1$yearID, RunIndex.LO.predict, lty="solid", col="black", lwd=1.5)
  lines(Team1$yearID, RunIndex.LO.25.predict, lty="dashed", col="red", lwd=2)
  lines(Team1$yearID, RunIndex.LO.5.predict, lty="dotdash", col="blue", lwd=1.5)
# chart legend and grid lines
  legend(firstyear+2, 140, 
    c("default", "span=0.25", "span=0.50"), 
    lty=c("solid", "dashed", "dotdash"), 
    col=c("black", "red", "blue"), 
    lwd=c(1.5, 2, 1.5))
  grid()
# straight line at 100
  abline(h = 100, lty="dotdash") 
#
#


```

The End
-30-