---
title: "Lahman Package - Issue Testing"
author: "Martin Monkman"
date: "August 29, 2015"
output: html_document
---

Testing issues in the Lahman package (version 3.0-1)


## Issue #7
Appearances Example Returns Bad Data 
[https://github.com/cdalzell/Lahman/issues/7](https://github.com/cdalzell/Lahman/issues/7)

### Using the Lahman package:

```{r}
library(Lahman)
library(plyr)

# Individuals who have played all 162 games since 1961
all162 <- ddply(subset(Appearances, yearID > 1960), .(yearID),
                         summarise, allGamers = playerID[G_all == 162])
# Number of all-gamers by year
table(all162$yearID)

```

### Using the original Lahman csv file

Hypothesis: this is a problem in the source data

#### 2013 version (i.e. same as Lahman package 3.0-1)

Notes: 
- the 'Appearances' table 2013 version of the Lahman database (csv format) was unzipped and renamed 'Appearances_2013'.
- the source file is lahman-csv_2014-02-14.zip, from [http://www.seanlahman.com/baseball-archive/statistics/] (http://www.seanlahman.com/baseball-archive/statistics/)

This csv file is thenis read into the R workspace and the same code is run.

```{r}
Appearances_2013 <- read.csv("Appearances_2013.csv")

all162_2013 <- ddply(subset(Appearances_2013, yearID > 1960), .(yearID),
                         summarise, allGamers = playerID[G_all == 162])
table(all162_2013$yearID)
```

#### 2014 version

Notes: 
- the 'Appearances' table 2014 version of the Lahman database (csv format) was unzipped and renamed 'Appearances_2014'.
- the source file is lahman-csv_2015-01-24.zip, from [http://www.seanlahman.com/baseball-archive/statistics/] (http://www.seanlahman.com/baseball-archive/statistics/)

This csv file is thenis read into the R workspace and the same code is run.

```{r}
Appearances_2014 <- read.csv("Appearances_2014.csv")

all162_2014 <- ddply(subset(Appearances_2014, yearID > 1960), .(yearID),
                         summarise, allGamers = playerID[G_all == 162])
table(all162_2014$yearID)
```

### CONCLUSION

The issue is in the source data.

-30-