---
title: "R Notebook"
output: html_notebook
---


A short set of instructions that starts with the "Teams" table in the Lahman database
and summarizes it for the Shiny app

Thus rather than having the app 
1. remove unnecessary records (rows) and fields (columns) and
2. run the calculations for the runs-per-game, runs-allowed-per-game, and indexed versions of those
the calculations are conducted here.

This will vastly improve the performance of the app.




```{r packages}

library(dplyr)
library(tidyr)
library(purrr)

library(Lahman)

```



CREATE LEAGUE SUMMARY TABLES

load the Lahman data table "Teams", filter


```{r}

Teams <- Lahman::Teams

# select a sub-set of teams from 1901 [the establishment of the American League] forward to most recent year
LG_year <- Teams %>%
  filter(yearID > 1900, lgID != "FL") %>%
  group_by(yearID, lgID) %>%
  nest()

LG_year

LG_year$data[[1]]

```

gapminder example

```{r}

library(gapminder)

by_country <- gapminder %>%
  group_by(country, continent) %>%
  nest()

by_country

by_country$data[[1]]

country_model <- function(df) {
  lm(lifeExp ~ year, data = df)
}

models <- purrr::map(by_country$data, country_model)

by_country2 <- by_country %>%
  mutate(model = map(data, country_model))

by_country2

```


## league run scoring function

a test

```{r}

leagueRuns_fun <- function(df) {
  sum(data = df$R)
}

league_year_runs <- map(LG_year$data, leagueRuns_fun)

league_year_runs[[1]]

Teams %>%
  filter(yearID == 1901,
         lgID == "AL") %>%
  summarise(leagueruns = sum(R))

```


the real mccoy

```{r}


leagueRPG_fun <- function(df) {
  sum(data = df$R) / sum(data = df$G)
  }

leagueRAPG_fun <- function(df) {
  sum(data = df$RA) / sum(data = df$G)
  }

league_year_RPG <- map(LG_year$data, leagueRPG_fun)

LG_year2 <- LG_year %>%
  mutate(lgRPG = map_dbl(LG_year$data, leagueRPG_fun),
         lgRAPG = map_dbl(LG_year$data, leagueRAPG_fun))

LG_year2

```



```{r}

#
#write.csv(LG_RPG, file="LG_RPG.csv", row.names=FALSE)
#
# and a version with just the MLB totals
MLB_RPG <- Teams %>%
  filter(yearID > 1900, lgID != "FL") %>%
  group_by(yearID) %>%
  summarise(R=sum(R), RA=sum(RA), G=sum(G)) %>%
  mutate(leagueRPG=R/G, leagueRAPG=RA/G)
#
#write.csv(MLB_RPG, file="MLB_RPG.csv", row.names=FALSE)
#
# create data frame Teams.merge with league averages
Teams.merge <-  Teams %>%
  mutate(teamRPG=(R/G), teamRAPG=(RA/G), WLpct=(W/G)) 
Teams.merge <- 
  merge(Teams.merge, LG_RPG, by = c("yearID", "lgID"))
#
# create new values to compare the individual team's runs/game compares to the league average that season
Teams.merge <- Teams.merge %>%
  # runs scored index where 100=the league average for that season
  mutate(R_index = (teamRPG/leagueRPG)*100) %>%
  mutate(R_index.sd = sd(R_index)) %>%
  mutate(R_z = (R_index - 100)/R_index.sd) %>%
  # runs allowed
  mutate(RA_index = (teamRAPG/leagueRAPG)*100) %>%
  mutate(RA_index.sd = sd(RA_index)) %>%
  mutate(RA_z = (RA_index - 100)/RA_index.sd)
#
# select only those columns which will be required
#   (port from shiny server file)

Teams.merge <- Teams.merge %>%
  select(yearID, lgID, franchID, teamID, name, 
         W, L, WLpct, R.x, RA.x, teamRPG, leagueRPG, 
         R_index, teamRAPG, leagueRAPG, RA_index)


#
#write.csv(Teams.merge, file="Teams_merge.csv", row.names=FALSE)

# -------------- THE END ...

```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
