---
title: "nested tables: Lahman example"
output: html_notebook
---




```{r packages}

# tidyverse packages
library(dplyr)
library(purrr)
library(readr)
library(tidyr)

library(feather)

#library(Lahman)


```


## gapminder example

From the "Many Models"" chapter of Wickham and Grolemund, [R for Data Science](https://r4ds.had.co.nz/many-models.html)


```{r}

library(gapminder)

gapminder

by_country <- gapminder %>%
  group_by(country, continent) %>%
  nest()

by_country

by_country$data[[1]]

# function to run regression model `lm()` on life expectancy by year
country_model <- function(df) {
  lm(lifeExp ~ year, data = df)
}

# single model object version
models <- purrr::map(by_country$data, country_model)

# nested table version
# - adds a new column "model" with model object
by_country <- by_country %>%
  mutate(model = map(data, country_model))

by_country

```



## MLB Lahman version

### Create league summary tables

A short set of instructions that starts with the "Teams" table in the Lahman database
and summarizes it for [MLB run scoring trends](https://monkmanmh.shinyapps.io/MLBrunscoring_shiny/) Shiny app

Thus rather than having the app do the work of

1. remove unnecessary records (rows) and fields (columns) and

2. run the calculations for the runs-per-game, runs-allowed-per-game, and indexed versions of those,

the calculations are conducted here. This will vastly improve the performance of the app.


#### Step 1

~~load the Lahman data table "Teams"~~

load the data table "Teams" (sourced from the baseballdatabank on github https://github.com/chadwickbureau/baseballdatabank))



- filter and 

- add team Runs Per Game (`RPG`), Runs Allowed Per Game (`RAPG`), and Win/Loss percentage (`WLpct`)

- nest, after `group_by()` year and league (`lgID`)


Then take a quick look at the first entry ... the American League, 1901

```{r}

#library(Lahman)

#Teams <- Lahman::Teams
Teams <- read_csv("Teams.csv")

# select a sub-set of teams from 1901 [the establishment of the American League] forward to most recent year
Teams_lgyr <- Teams %>%
  filter(yearID > 1900, lgID != "FL") %>%
  group_by(yearID, lgID) %>%
  nest()

Teams_lgyr

Teams_lgyr$data[[1]]


```



#### Step 2

create a league run scoring function

**Note:**

- In the gapminder example, the variables were called using their names. In this case, for a reason I have not yet determined, we have to specify the data object they are coming from; e.g. for the runs variable `R`, we have to use `df$R` (not just `R`). 


First, a simple test:

```{r}

# base R format
leagueRuns_fun <- function(df) {
  sum(data = df$R)
}

# dplyr format
#leagueRuns_fun <- function(df) {
#  df %>% summarize(sum=sum(R))
#}

league_year_runs <- map(Teams_lgyr$data, leagueRuns_fun)

league_year_runs[[1]]

#check the answer by old school `dplyr` method
Teams %>%
  filter(yearID == 1901,
         lgID == "AL") %>%
  summarise(leagueruns = sum(R))

```


now for the real thing

**Note:**

- There are two functions, one for Runs and the other for Runs Allowed. This is because I have not yet figured out how to specify two different variables (i.e. the name of the data object and the variable to be used in the function) in the map_() function and successfully have them carried into my calculation functions


1st approach: put the sum calculation inside the nested tibble


Also note that in order to be consistent with other sources, the number of games played is calculated using the sum of wins (`W`) and losses (`L`), rather than the number of games reported in the `G` variable. 


```{r}

# functions
leagueRPG_fun <- function(df) {
  sum(data = df$R) / (sum(data = df$W) + sum(data = df$L))
  }

leagueRAPG_fun <- function(df) {
  sum(data = df$RA) / (sum(data = df$W) + sum(data = df$L))
  }


# simple `map` version
league_year_RPG <- map(Teams_lgyr$data, leagueRPG_fun)

# embed as new columns in nested data object
Teams_lgyr <- Teams_lgyr %>%
  mutate(lgRPG = map_dbl(Teams_lgyr$data, leagueRPG_fun),
         lgRAPG = map_dbl(Teams_lgyr$data, leagueRAPG_fun))

Teams_lgyr



```


2nd approach: sum league total runs, runs allowed, and games, then RPG and RAPG outside the nested tibbles

```{r}

# another function - calculate number fo games played
leagueG_fun <- function(df) {
  (sum(data = df$W) + sum(data = df$L))
}


Teams_lgyr <- Teams_lgyr %>%
  mutate(lgR = map_dbl(Teams_lgyr$data, leagueR_fun),
         lgRA = map_dbl(Teams_lgyr$data, leagueRA_fun),
         lgG = map_dbl(Teams_lgyr$data, leagueG_fun))


Teams_lgyr <- Teams_lgyr %>%
  mutate(lgRPG = (lgR / lgG),
         lgRAPG = (lgRA / lgG))

Teams_lgyr


```



#### save LG_RPG files


Save feather and csv versions for shiny app

- note renaming of variables to correspond with shiny app names

```{r}

LG_RPG <- Teams_lgyr %>%
  select(yearID, lgID, R = lgR, RA = lgRA, G = lgG, 
         leagueRPG = lgRPG, leagueRAPG = lgRAPG)

write_csv(LG_RPG, "LG_RPG.csv")
write_feather(LG_RPG, "LG_RPG.feather")

```


### repeat for MLB total

This only differs in the level of nesting; instead of grouping by year and league, it's only year  (`yearID`).

```{r}

Teams_lgyr <- Teams_lgyr %>%
  unnest() %>%
  group_by(yearID) %>%
  nest()

Teams_lgyr

Teams_lgyr <- Teams_lgyr %>%
  mutate(mlbR = map_dbl(Teams_lgyr$data, leagueR_fun),
         mlbRA = map_dbl(Teams_lgyr$data, leagueRA_fun),
         mlbG = map_dbl(Teams_lgyr$data, leagueG_fun),
         mlbRPG = (mlbR / mlbG),
         mlbRAPG = (mlbRA / mlbG))

Teams_lgyr

```


Save feather and csv versions for shiny app

- note renaming of variables to correspond with shiny app names

```{r}

MLB_RPG <- Teams_lgyr %>%
  select(yearID, R = mlbR, RA = mlbRA, G = mlbG, 
         leagueRPG = mlbRPG, leagueRAPG = mlbRAPG)

write_csv(MLB_RPG, "MLB_RPG.csv")
write_feather(MLB_RPG, "MLB_RPG.feather")

```



### Step 3

Calculate index of team run scoring against league average 

Note that we start with `unnest()` and create a new object, `Teams_append` ... a tibble with all of the variables exposed.


```{r}

Teams_append <- Teams_lgyr %>%
  unnest() %>%
  mutate(teamRPG=(R / (W + L)), 
         teamRAPG=(RA / (W + L)), 
         WLpct=(W / (W + L))) %>%
  # runs scored index where 100=the league average for that season
  mutate(R_index = (teamRPG / lgRPG) * 100) %>%
  mutate(R_index.sd = sd(R_index)) %>%
  mutate(R_z = (R_index - 100) / R_index.sd) %>%
  # runs allowed
  mutate(RA_index = (teamRAPG / lgRAPG) * 100) %>%
  mutate(RA_index.sd = sd(RA_index)) %>%
  mutate(RA_z = (RA_index - 100) / RA_index.sd)
#

Teams_append

```



Save feather and csv versions for shiny app

- note selection and renaming of variables to correspond with shiny app names

```{r}

Teams_merge <- Teams_append %>%
  select(yearID, lgID, franchID, teamID, name,
         W, L, WLpct, R.x = R, RA.x = RA, 
         teamRPG, leagueRPG = lgRPG, R_index,
         teamRAPG, leagueRAPG = lgRAPG, RA_index)

Teams_merge

write_csv(MLB_RPG, "MLB_RPG.csv")
write_feather(MLB_RPG, "MLB_RPG.feather")

```



-30-

