---
title: "MLB run scoring trends: data wrangle"
author: "Martin Monkman"
date: "February 19, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




```{r packages}

# tidyverse packages
library(dplyr)
library(purrr)
library(readr)
library(tidyr)

library(feather)

#library(Lahman)


```


The new Major League Baseball season will soon begin, which means it's time to look back and update my run scoring trends data visualization application, built using RStudio's shiny package.

This update gave me the opportunity to make some cosmetic tweaks to the front end, but more consequential changes to the code under the hood.

### 1. retired `reshape`, using `tidyr`

At one point in the app's code, I had used the now-retired [`reshape`](https://cran.r-project.org/web/packages/reshape/index.html) package to melt a data table. Although this still works, I opted to update the code to use the `gather()` function in the package `tidyr`. 

### 2. `feather` instead of csv

The app relied on some pre-wrangled csv files; these have been replaced by files stored using the [.feather format](https://blog.rstudio.com/2016/03/29/feather/), which makes for a signficant performance improvement. 

Here's a speed test comparison of the largest file (by both columns and rows), using my home PC:

```{r}

test1 <- system.time(
  dat1 <- read.csv("Teams_merge.csv")
)
test1

test2 <- system.time(
  dat2 <- read_csv("Teams_merge.csv")
)
test2

test3 <- system.time(
  dat3 <- read_feather("Teams_merge.feather")
)
test3

```


### 3. Wrangling: the calculation of team and league run scoring averages

The goal is to create data tables that minimize the amount of processing the app has to do. 

In previous versions of the app, the filtering of rows (records or observations) and selecting of columns (variables), the calculation of each team's average runs scored and runs allowed per game, the league average runs per game, and the comparison of the team to the league, was done first using base R's `apply` family of functions, along with XYZ

Then I switched to using `dplyr`, and although the steps were now in a pipe, it still required creating a separate data table with the league average, and then joining that table back into the main team table so that the team result could be compared to the league average.

For this iteration, preparing the data for the app is now done using `tidyr::nest()` and `purrr::map()`. What follows is a detailed explanation of how I approached this. 

It's always valuable to have your end-state in mind when working through a multi-step data wrangle like this. My goal as the values shown on the "team plot" tab of the app -- an index value (i.e. a percentage) of a team's average runs scored (and runs allowed) compared to the league run scoring rate, for a single season.



#### Read the data

The first step is to read in the data.

- It comes from the "Lahman", so-named after the originator and hoster, Sean Lahman. [His website](http://www.seanlahman.com/baseball-archive/statistics/) has the current version of the database, as well as an archive of previous versions.

- The data table "Teams" is sourced from the _baseballdatabank_ on github https://github.com/chadwickbureau/baseballdatabank))

- There is an R package with the Lahman database, but it's currently two seasons behind the source files.


```{r}
#Teams <- Lahman::Teams
Teams <- read_csv("Teams.csv")

Teams

```


The table above has far more variables than what we need, and some that we'll have to calculate. 



### Create league summary tables

A short set of instructions that starts with the "Teams" table in the Lahman database
and summarizes it for [MLB run scoring trends](https://monkmanmh.shinyapps.io/MLBrunscoring_shiny/) Shiny app

Thus rather than having the app do the work of

1. remove unnecessary records (rows) and fields (columns) and

2. run the calculations for the runs-per-game, runs-allowed-per-game, and indexed versions of those,

the calculations are conducted here. This will vastly improve the performance of the app.


#### Step 1


I started with the "Many Models"" chapter of Wickham and Grolemund, [R for Data Science](https://r4ds.had.co.nz/many-models.html). (And thanks to Dr. Charlotte Wickham, whose training course was invaliable in helping me wrap my head around this.)

At this point, the code 

- filters out the years prior to 1901 and [the misbegotten Federal League](https://en.wikipedia.org/wiki/Federal_League).

- and then creates a nested data table, starting with the `group_by()` year and league (`lgID`)


```{r}

# select a sub-set of teams from 1901 [the establishment of the American League] forward to most recent year
Teams_lgyr <- Teams %>%
  filter(yearID > 1900, lgID != "FL") %>%
  group_by(yearID, lgID) %>%
  nest()

Teams_lgyr

```


Here's a quick peek inside the first entry of the "data" column...the American League, 1901.

```{r}

Teams_lgyr$data[[1]]


```



#### Step 2

This step creates a league run scoring function, and then applies that using the `purrr::map()` function.

**Note:**

- In the gapminder example, the variables were called using their names. In this case, for a reason I have not yet determined, we have to specify the data object they are coming from; e.g. for the runs variable `R`, we have to use `df$R` (not just `R`). 


First, a simple test:

```{r}

# base R format
leagueRuns_fun <- function(df) {
  sum(data = df$R)
}

# dplyr format
#leagueRuns_fun <- function(df) {
#  df %>% summarize(sum=sum(R))
#}

league_year_runs <- map(Teams_lgyr$data, leagueRuns_fun)

league_year_runs[[1]]

#check the answer by old school `dplyr` method
Teams %>%
  filter(yearID == 1901,
         lgID == "AL") %>%
  summarise(leagueruns = sum(R))

```


now for the real thing

**Note:**

- There are two functions, one for Runs and the other for Runs Allowed. This is because I have not yet figured out how to specify two different variables (i.e. the name of the data object and the variable to be used in the function) in the map_() function and successfully have them carried into my calculation functions


1st approach: put the sum calculation inside the nested tibble


Also note that in order to be consistent with other sources, the number of games played is calculated using the sum of wins (`W`) and losses (`L`), rather than the number of games reported in the `G` variable. 


```{r}

# functions
leagueRPG_fun <- function(df) {
  sum(data = df$R) / (sum(data = df$W) + sum(data = df$L))
  }

leagueRAPG_fun <- function(df) {
  sum(data = df$RA) / (sum(data = df$W) + sum(data = df$L))
  }


# simple `map` version
league_year_RPG <- map(Teams_lgyr$data, leagueRPG_fun)

# embed as new columns in nested data object
Teams_lgyr <- Teams_lgyr %>%
  mutate(lgRPG = map_dbl(Teams_lgyr$data, leagueRPG_fun),
         lgRAPG = map_dbl(Teams_lgyr$data, leagueRAPG_fun))

Teams_lgyr



```


2nd approach: sum league total runs, runs allowed, and games, then RPG and RAPG outside the nested tibbles

```{r}

# more functions - individual league by year totals
leagueR_fun <- function(df) {
  sum(data = df$R)
}

leagueRA_fun <- function(df) {
  sum(data = df$RA)
}

leagueG_fun <- function(df) {
  (sum(data = df$W) + sum(data = df$L))
}


Teams_lgyr <- Teams_lgyr %>%
  mutate(lgR = map_dbl(Teams_lgyr$data, leagueR_fun),
         lgRA = map_dbl(Teams_lgyr$data, leagueRA_fun),
         lgG = map_dbl(Teams_lgyr$data, leagueG_fun))


Teams_lgyr <- Teams_lgyr %>%
  mutate(lgRPG = (lgR / lgG),
         lgRAPG = (lgRA / lgG))

Teams_lgyr


```



#### save LG_RPG files


Save feather and csv versions for shiny app

- note renaming of variables to correspond with shiny app names

```{r}

LG_RPG <- Teams_lgyr %>%
  select(yearID, lgID, R = lgR, RA = lgRA, G = lgG, 
         leagueRPG = lgRPG, leagueRAPG = lgRAPG)

LG_RPG

write_csv(LG_RPG, "LG_RPG.csv")
write_feather(LG_RPG, "LG_RPG.feather")

```


### repeat for MLB total

This only differs in the level of nesting; instead of grouping by year and league, it's only year  (`yearID`).

```{r}

Teams_lgyr <- Teams_lgyr %>%
  unnest() %>%
  group_by(yearID) %>%
  nest()

Teams_lgyr

Teams_lgyr <- Teams_lgyr %>%
  mutate(mlbR = map_dbl(Teams_lgyr$data, leagueR_fun),
         mlbRA = map_dbl(Teams_lgyr$data, leagueRA_fun),
         mlbG = map_dbl(Teams_lgyr$data, leagueG_fun),
         mlbRPG = (mlbR / mlbG),
         mlbRAPG = (mlbRA / mlbG))

Teams_lgyr

```


Save feather and csv versions for shiny app

- note renaming of variables to correspond with shiny app names

```{r}

MLB_RPG <- Teams_lgyr %>%
  select(yearID, R = mlbR, RA = mlbRA, G = mlbG, 
         leagueRPG = mlbRPG, leagueRAPG = mlbRAPG)

write_csv(MLB_RPG, "MLB_RPG.csv")
write_feather(MLB_RPG, "MLB_RPG.feather")

```



### Step 3

Calculate index of team run scoring against league average 

Note that we start with `unnest()` and create a new object, `Teams_append` ... a tibble with all of the variables exposed.


```{r}

Teams_append <- Teams_lgyr %>%
  unnest() %>%
  mutate(teamRPG=(R / (W + L)), 
         teamRAPG=(RA / (W + L)), 
         WLpct=(W / (W + L))) %>%
  # runs scored index where 100=the league average for that season
  mutate(R_index = (teamRPG / lgRPG) * 100) %>%
  mutate(R_index.sd = sd(R_index)) %>%
  mutate(R_z = (R_index - 100) / R_index.sd) %>%
  # runs allowed
  mutate(RA_index = (teamRAPG / lgRAPG) * 100) %>%
  mutate(RA_index.sd = sd(RA_index)) %>%
  mutate(RA_z = (RA_index - 100) / RA_index.sd)
#

Teams_append

```

Save feather and csv versions for shiny app

- note selection and renaming of variables to correspond with shiny app names

```{r}

Teams_merge <- Teams_append %>%
  select(yearID, lgID, franchID, teamID, name,
         W, L, WLpct, R.x = R, RA.x = RA, 
         teamRPG, leagueRPG = lgRPG, R_index,
         teamRAPG, leagueRAPG = lgRAPG, RA_index)

Teams_merge

write_csv(Teams_merge, "Teams_merge.csv")
write_feather(Teams_merge, "Teams_merge.feather")

```



-30-

