---
title: "NewCellPhonePuzzle"
author: "Martin Monkman"
date: "December 28, 2015"
output: html_document
---

# Solving a Puzzle

FiveThirtyEight regularly runs a feature called [The Riddler](http://fivethirtyeight.com/tag/the-riddler/). The puzzle from December 22, 2015 was "[How Long Will Your Smartphone Distract You From Family Dinner?]"(<http://fivethirtyeight.com/features/how-long-will-your-smartphone-distract-you-from-family-dinner/>):

> You've just finished unwrapping your holiday presents. You and your sister got brand-new smartphones, opening them at the same moment. You immediately both start doing important tasks on the Internet, and each task you do takes one to five minutes. (All tasks take exactly one, two, three, four or five minutes, with an equal probability of each). After each task, you have a brief moment of clarity. During these, you remember that you and your sister are supposed to join the rest of the family for dinner and that you promised each other you'd arrive together. You ask if your sister is ready to eat, but if she is still in the middle of a task, she asks for time to finish it. In that case, you now have time to kill, so you start a new task (again, it will take one, two, three, four or five minutes, exactly, with an equal probability of each). If she asks you if it's time for dinner while you're still busy, you ask for time to finish up and she starts a new task and so on. From the moment you first open your gifts, how long on average does it take for both of you to be between tasks at the same time so you can finally eat? (You can assume the "moments of clarity" are so brief as to take no measurable time at all.)

To solve this, I decided to take (as is my wont) a Monte Carlo type approach, running repeated trails.

### A single scenario

The first step was to write some R code that run through the scenario once.  In this, I used the `sample` function, which takes a random sample from a vector (in this case, 1 to 5) a specified number of times.  First, we take two samples, and assign them to `task_a` (for the larger of the two values) and `task_b` (for the smaller).

Then the code moves into a `while` loop, running as long as the values of `task_a` and `task_b` are unequal.  If they are unequal, a sample of 1 is drawn and that value is added to `task_b`, and the loop starts again with the comparison of the two values.

Once they are equal, the loop breaks and the message "Dinner time!" along with the elapsed time is printed.

```{r}

# #-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-

firsttask <- sample(1:5, 2, replace=TRUE)
firsttask

task_a <- max(firsttask)
task_a
  
task_b <- min(firsttask)
task_b

while (TRUE) {
  
  if (task_a == task_b) break

  if (task_a != task_b)
  print("keep tasking")

  print(paste("task_a = ", task_a))
  print(paste("task_b  = ", task_b))
  
  nexttask <- sample(1:5, 1, replace=TRUE)
  print(paste("nexttask = ", nexttask))

  task_b <- nexttask + task_b

  task_min <- min(task_a, task_b)

  task_max <- max(task_a, task_b)

  task_a <- task_max
  task_b <- task_min

}

print("Dinner time!")
print(paste("It took ", task_a, " minutes"))




```

### Multiple Runs

Running the scenario once doesn't tell us what we need to know. To get there we need a volume of samples to approach the mean.

So we turn the code (without all the printing along the way) above into a function called `dinnerTime_fn`.


```{r}

dinnerTime_fn <- function() 
  {
  firsttask <- sample(1:5, 2, replace=TRUE)

  task_a <- max(firsttask)
  task_b <- min(firsttask)

  while (TRUE) {
  
    if (task_a == task_b) break

    if (task_a != task_b)
      nexttask <- sample(1:5, 1, replace=TRUE)
      task_b <- nexttask + task_b
      task_min <- min(task_a, task_b)
      task_max <- max(task_a, task_b)
      
      task_a <- task_max
      task_b <- task_min
  }

return(task_a)
}


```

A simple call of the function now runs all of the above code and returns the length of time to get to dinner.

```{R}

dinnerTime_fn()

```

With the function now defined, we can use the instruction `replicate` to create a new vector with multiple return values of `dinnerTime`.  And from the vector of values, we can calculate the mean and median along with the minimum and maximum, and plot a quick and dirty histogram with `hist`.


```{r}


multipledinnerTimes <- replicate(10000, dinnerTime_fn())

# summarizing multipledinnerTimes
summary(multipledinnerTimes)
hist(multipledinnerTimes)

                 
```

So that was easy.  But it could be compounded--if we were to run `multipledinnerTimes` over and over, we'd see some degree of variation. (This is true even with batches each with 10,000 samples; remember that this is the uniform distribution. The normal distribution has higher probabilities of the values closer to the mean, and thus takes fewer iterations to home in on the theoretical value.)


### Monte Carloing the Monte Carlo

```{r}

multipledinnerTimes_fn <- function() {
  multipledinnerTimes <- replicate(10000, dinnerTime_fn())
  return(mean(multipledinnerTimes))
}

# a single iteration of the multiple runs function
multipledinnerTimes_fn()

# and multiple multiples
multipleMultiples <- replicate(100, multipledinnerTimes_fn())
# summarizing multipledinnerTimes
summary(multipleMultiples)
hist(multipleMultiples)


```

## The Actual Solution

The [theortical solution is here (scroll below the puzzle)](http://fivethirtyeight.com/features/riddler-how-long-before-you-can-use-your-2015-calendar-again-riddler/) (spoiler: 9 minutes), along with another approach in R.


-30-
